---
title: "Find bee SRA runs from metadata"
author: "Blake Paget"
date: '2025'
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background
RNA-Seq data will be analysed to determine the expression pattern of an uncharacterised protein.
The Sequence Read Archive (SRA) maintained by NCBI will be used. The metadata will be searched for appropriate runs. In a separate publication selected runs will be downloaded and analysed.<br>

"SRA has deposited its metadata into BigQuery to provide the bioinformatics community with programmatic access to this data.
You can now search across the entire SRA by sequencing methodologies and sample attributes."<br>

A simple SQL query will be used to retrieve a table for further analysis in R. The sample attributes are an inconsistent mess. R will be used to filter the table.

## Script
#### Load libraries
```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(bigrquery)
library(dotenv)
library(ggplot2)
```
<br>

#### Import csv from BigQuery
```{r bq}
# Use environment variables to mask private data.
load_dot_env()

# Read the csv bq table if available, otherwise redo the SQL query.
df <- try(read.csv("bquxjob_722b10fc_1971e82ea0f.csv"))
if (inherits(df, "try-error")) {
  # BigQeury SQL query
  query <-
    "
    SELECT *
    FROM `nih-sra-datastore.sra.metadata`
    WHERE
      organism = 'Apis mellifera' AND
      consent = 'public' AND
      assay_type = 'RNA-Seq'
    "
  tb <- bq_project_query(Sys.getenv("BIGQUERY_PROJECT"), query)
  df <- bq_table_download(tb)
}
```
<br>

#### Data cleaning
```{r cleaning1}
glimpse(df)

# Any NA rows for primary key 'acc'?
is.na(df$acc) %>% table()

# Check file sizes.
is.na(df$mbytes) %>% table()
max(df$mbytes)
min(df$mbytes)

# What entries have 0 Mb?
df %>%
  filter(mbytes == 0) %>% 
  group_by(bioproject) %>% 
  count()

df %>% 
  filter(bioproject == "PRJNA477521") %>% 
  select(acc, mbases, mbytes, library_name) %>% 
  glimpse

# One bioproject with samples with blank reads? Remove from list.
df <-
  df %>% 
  filter(mbytes != 0)

nrow(df)

# Looks like the 'attributes' and 'jattr' are nested as JSON objects. 
# Not shown here, I have extracted these and compared them.
# These each contain the same sample info, e.g. 'tissue_sam_ss_dpl145': 'thoracic muscle'.
# Expand 'jattr' and add it to the df. Drop 'jattr' and 'attributes' nested columns.
```
```{r cleaning2, warning=FALSE, message=FALSE}
library(jsonlite)
```
```{r cleaning3}
parsed_jattr <- lapply(df$jattr, fromJSON)
jattr_tibble <- 
  lapply(parsed_jattr, function(x) { 
    t(cbind(names(x), as.character(unname(x))))
  }) %>% 
  lapply(function(x) {
    z <- as.data.frame(x[-1, ])
    row.names(z) <- x[1, ]
    as.data.frame(t(z))
  }) %>% 
  do.call(plyr::rbind.fill, .) %>% 
  as_tibble

df_wide <- as_tibble(cbind(df[, !names(df) %in% c("attributes", "jattr")], jattr_tibble), .name_repair = "unique")
```
<br>

#### Data analysis

The RNA-seq data will be sectioned by caste and large body structure initially. Only adult bees produce honey or royal jelly therefore earlier developmental stages will be excluded. The keywords are listed in the code chunk below.<br>


```{r analysis1}
```

```{r analysis2}
# Table occurrences of keywords by column.

# Define keywords and exclusions. 
# Note: bounded words \\b;
#       plural forms included;
#       "whole.?body" is used so regex searches for any character in between the two words 0 or 1 time.
keywords <- c("abdomen", "drone", "female", "forager", "head", "male", "nurse", "queen", "thorax", "whole body", "worker")
keywords_re <- keywords
keywords_re[-c(3, 6, 10)] <- paste0("\\b", keywords[-c(3, 6, 10)], "(s)?\\b")
keywords_re[c(3, 6)] <- paste0("\\b", keywords[c(3,6)], "\\b")
keywords_re[c(9,10)] <- c("\\bthora[xc](es)?\\b", "\\bwhole.?bod[yi](es)?\\b")

exclusions_re <- 
  c("embryo(s)?", "larva(e)?", "pupa(e)?") %>% 
  paste0("\\b", ., "\\b")

# Exclude rows from the expanded column
rows_to_exclude <- list()
for (exclude in exclusions_re) {
  rows_to_exclude[[exclude]] <-
    df_wide %>% 
    select(where(is.character)) %>% 
    sapply(., function(x) { grep(exclude, x, ignore.case = T)}) %>%
    unlist %>% 
    as.vector %>% 
    unique
}
rows_to_exclude <- 
  rows_to_exclude %>% 
  unlist %>% 
  unique
df_wide_ex <- df_wide[-rows_to_exclude, ]

# Grep each row of the table for the keyword, producing a list of all columns
# with the row number under the column the keyword appeared in.
# Count the number of row indices for each column (length).
keyword_list <- list()
for (keyword in keywords_re) {
  keyword_list[[keyword]] <- 
    df_wide_ex %>% 
    select(where(is.character)) %>%
    sapply(., function(x) { grep(keyword, x, ignore.case = T)}) %>%
    lapply(., length) %>% 
    unlist
}

# Tidy.
keyword_find <- as.data.frame(do.call(rbind, keyword_list))
keyword_find <- keyword_find[ , colSums(keyword_find) != 0]
ncol(keyword_find) # the keywords appear in this many columns of the expanded metadata.
```
<br>

The table below gives the number of times any keyword appears in a column.

```{r analysis3}
# Summary of keyword spread.
data.frame(freq = sort(colSums(keyword_find), decreasing = T)) # most frequent columns
```
<br>

SRA runs that are clearly defined by caste:tissue or sex:tissue will be summarised and examined further. Two keyword combinations will be used to search across all character columns. Number of SRA runs and the total file size can then be calculated.<br>

```{r analysis4}
# All combinations (x2) of keywords with replacement and distinct items.
keywords_x2 <- arrangements::combinations(x = keywords_re, k = 2, replace = TRUE)

# Filter for both keywords, ignore case.
df_wide_in <- apply(keywords_x2, 1, function(x) {
  df_wide_ex %>% 
    select(where(is.character)) %>% 
    filter(if_any(everything(), ~ grepl(x[1], ., ignore.case = T)) & if_any(everything(), ~ grepl(x[2], ., ignore.case = T)))
})
df_wide_in_count <- lapply(df_wide_in, count)

# Combine keyword combinations and their counts.
keywords_x2_count <-
  as.data.frame(cbind(
    arrangements::combinations(x = keywords, k = 2, replace = TRUE),
    unlist(df_wide_in_count)))
```
<br>

Compute power is limited so file size is a big consideration.

```{r analysis5}
# df_wide_in contains only the chr columns, need 'mbytes' column from df_wide
df_wide_sizes <-
  df_wide_in %>% 
  lapply(., select, acc) %>% 
  lapply(., function(x) {
    df_wide %>% 
      semi_join(x, by = "acc") %>% 
      summarise(sum = sum(mbytes)) %>% 
      pull(sum)
  })


# Add gb info to key_key_count table. 
keywords_x2_count$gb <- round(unlist(df_wide_sizes)/1000, digits = 0)

# Duplicate the counts for the reciprocal combinations to make the plot more readable.
recip_key_x2_count <- 
  rename(keywords_x2_count, V1 = V2, V2 = V1) %>% 
  bind_rows(., keywords_x2_count) %>% 
  distinct
  
recip_key_x2_count$V3 <- as.numeric(recip_key_x2_count$V3)
recip_key_x2_count$gb <- as.numeric(recip_key_x2_count$gb)
```
<br>

#### Plot

```{r analysis6}
ggplot(recip_key_x2_count, aes(V1, V2)) +
  geom_tile(aes(fill = V3)) +
  geom_text(aes(label = V3)) +
  scale_fill_gradient(low = "white",
                      high = "orange",
                      name = "SRA runs") +
  labs(title = "Number of SRA runs given a combination of two keywords",
       subtitle = "Diagonal represents a single keyword") +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4, size = 12),
        axis.text.y = element_text(size = 12),
        axis.line.x = element_line(color = "black"),
        axis.line.y = element_line(color = "black"),
        panel.grid.major = element_blank(),
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0),
        plot.title.position = "plot",
        panel.background = element_blank(),
        panel.border = element_rect(color = "grey60", fill = NA),
        legend.position = "right",
        strip.background = element_rect(color = "grey60", fill = "white"))
```
<br>

Total SRA run file sizes. I assume original file size, not SRAlite file size.

```{r analysis7}
ggplot(recip_key_x2_count, aes(V1, V2)) +
  geom_tile(aes(fill = gb)) +
  geom_text(aes(label = gb)) +
  scale_fill_gradient(low = "white",
                      high = "lightblue",
                      name = "Size (Gb)") +
  labs(title = "Total size of SRA runs given a combination of two keywords",
       subtitle = "Diagonal represents a single keyword") +
  theme(axis.title = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.4, size = 12),
        axis.text.y = element_text(size = 12),
        axis.line.x = element_line(color = "black"),
        axis.line.y = element_line(color = "black"),
        panel.grid.major = element_blank(),
        plot.caption.position = "plot",
        plot.caption = element_text(hjust = 0),
        plot.title.position = "plot",
        panel.background = element_blank(),
        panel.border = element_rect(color = "grey60", fill = NA),
        legend.position = "right",
        strip.background = element_rect(color = "grey60", fill = "white"))
```
<br>